---
title: "MSL2 Immunofluorescence Image Feature Analysis"
author: "Muhunden Jayakrishnan / Fotios Gkountromichos"
date: "2025"
output:
  html_document:
    fig_caption: true
---

# Figures produced
# - Fig 2E:   Combined UMAP coloured by genotype + individual genotype UMAPs
# - SFig 1A:  Replicate UMAPs (run this script on each replicate directory)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!dir.exists("figures")) dir.create("figures")

library(tiff)
library(tidyverse)
library(ggplot2)
library(caret)
library(tidymodels)
library(corrr)
library(ggfortify)
library(GLCMTextures)
library(EBImage)
library(reticulate)
library(ComplexHeatmap)
library(umap)
library(cluster)
```

## Genotype extraction and colour palette

```{r genotype_helpers, echo=TRUE}
extract_genotype <- function(filename) {
  fname <- tolower(filename)
  if (grepl("rox2ko17_a0b0|a0b0", fname))             return("rox2ko17_a0b0")
  if (grepl("rox2ko17_a3b0|a3b0", fname))             return("rox2ko17_a3b0")
  if (grepl("minirox", fname))                          return("rox2ko17_minirox")
  if (grepl("rox2ko17_rox1d3f|rox1d3f", fname))       return("rox2ko17_rox1d3f")
  if (grepl("rox2ko17_rox2fl|rox2fl|rox2_fl", fname)) return("rox2ko17_rox2fl")
  if (grepl("rox2ko17", fname))                         return("rox2ko17")
  if (grepl("rox2ko112", fname))                        return("rox2ko112")
  if (grepl("wt", fname))                               return("wt")
  return(NA)
}

genotype_order <- c(
  "wt", "rox2ko112", "rox2ko17", "rox2ko17_rox2fl",
  "rox2ko17_minirox", "rox2ko17_a0b0", "rox2ko17_a3b0",
  "rox2ko17_rox1d3f"
)

genotype_colors <- c(
  wt               = "#009E73",
  rox2ko112        = "yellow3",
  rox2ko17         = "#f77f00",
  rox2ko17_rox2fl  = "springgreen3",
  rox2ko17_minirox = "#8338ec",
  rox2ko17_a0b0    = "#ff6600",
  rox2ko17_a3b0    = "#ff006e",
  rox2ko17_rox1d3f = "#3a86ff"
)
```

## Read images

Read all TIFF files once, extract MSL2 channel (channel 2), and filter to
multi-channel files only.

```{r read_images, echo=TRUE}
list_files <- list.files(path = "tiffs", pattern = "\\.tif$", full.names = TRUE)
image_genotypes <- sapply(list_files, extract_genotype)

cat("Reading TIFF files (this may take a moment)...\n")
image_data_list <- lapply(seq_along(list_files), function(i) {
  if (i %% 500 == 0) cat("  Processed", i, "of", length(list_files), "files...\n")
  img_channels <- suppressWarnings(readTIFF(list_files[i], as.is = TRUE, all = TRUE))
  list(
    is_multichannel = is.list(img_channels) && length(img_channels) >= 2,
    channel2 = if (is.list(img_channels) && length(img_channels) >= 2)
                 img_channels[[2]] else NULL
  )
})

is_multichannel <- sapply(image_data_list, function(x) x$is_multichannel)
cat("Total files found:", length(list_files), "\n")
cat("Multichannel files:", sum(is_multichannel), "\n")
cat("Single-channel files excluded:", sum(!is_multichannel), "\n")

list_files      <- list_files[is_multichannel]
image_genotypes <- image_genotypes[is_multichannel]
image_raw       <- lapply(image_data_list[is_multichannel], function(x) x$channel2)
unique_genotypes <- unique(image_genotypes)
rm(image_data_list)

cat("Genotypes after filtering:", paste(unique_genotypes, collapse = ", "), "\n")
```

## Display example nuclei

```{r example_nuclei, echo=TRUE, fig.height=8, fig.width=12}
set.seed(1254)
par(mfrow = c(2, 3))
for (i in seq_along(unique_genotypes)) {
  genotype_indices <- which(image_genotypes == unique_genotypes[i])
  if (length(genotype_indices) >= 6) {
    selected_indices <- genotype_indices[sample.int(length(genotype_indices), 6)]
  } else {
    selected_indices <- genotype_indices
  }
  for (j in seq_along(selected_indices)) {
    image(image_raw[[selected_indices[j]]], col = gray.colors(256),
          main = paste(unique_genotypes[i], "_", j), asp = 1)
    axis(1, 0:1, 0:1)
  }
}
```

## Per-cell mean intensity violin plot

```{r per_cell_violin, echo=TRUE, fig.height=5, fig.width=7, message=FALSE, warning=FALSE}
cell_stats <- data.frame(genotype = character(), mean_intensity = numeric())
for (i in seq_along(list_files)) {
  img <- image_raw[[i]]
  if (is.matrix(img)) {
    cell_stats <- rbind(cell_stats, data.frame(
      genotype       = image_genotypes[i],
      mean_intensity = mean(img)
    ))
  }
}

p <- ggplot(cell_stats, aes(x = genotype, y = mean_intensity, fill = genotype)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width = 0.1, outlier.size = 0.5, alpha = 0.7) +
  labs(title = "Per-Cell Mean MSL2 Intensity by Genotype",
       x = "Genotype", y = "Mean MSL2 Intensity") +
  theme_minimal()
print(p)
ggsave("figures/per_cell_mean_intensity_violin.png", p, width = 7, height = 5)
```

## Extract GLCM and object features (46 per cell)

Images are normalised to 0-1 range per image (`y / max(y)`) prior to
feature extraction (Methods: "Images were normalized to a 0-1 intensity
range prior to feature extraction").

```{r glcm_object_features, echo=TRUE, fig.keep='none'}
res_method <- as.data.frame(matrix(nrow = length(list_files), ncol = 46))

for (i in seq_along(list_files)) {
  y <- Image(image_raw[[i]])
  y <- y / max(y)

  imagename <- gsub("tiffs/MSL2gp_*", "", gsub(".tif", "", list_files[i]))

  if (i %in% c(19, 16, 28)) {
    plot(y)
    hist(as.vector(imageData(y)), main = imagename, breaks = 50)
  }

  # Haralick texture features (whole-image mask) -- 26 features
  x <- Image(1, dim = dim(y))
  res_method[i, 1:26] <- colMeans(computeFeatures.haralick(x, y))

  # Mean intensity -- 1 feature
  res_method[i, 27] <- mean(imageData(y))

  # Object detection: Gaussian blur + threshold at 0.85
  y_2 <- gblur(y, sigma = 0.3)
  cutoff <- 0.85
  x <- y_2 > cutoff

  if (i %in% c(19, 16, 28)) plot(x)

  # Object features (moment + shape + basic) -- 19 features
  if (sum(imageData(x)) == 0) {
    res_method[i, 28:46] <- 0
  } else {
    res_method[i, 28:46] <- c(
      computeFeatures.moment(x, y),
      computeFeatures.shape(x),
      computeFeatures.basic(x, y)
    )
  }
  rownames(res_method)[i] <- imagename
}
head(res_method)
```

## Remove duplicates

```{r remove_duplicates, echo=TRUE}
cat("Duplicated rows:", sum(duplicated(res_method)), "\n")
res_method <- res_method[!duplicated(res_method), ]
```

## Correlation plot

```{r correlation_plot, echo=TRUE, fig.height=8, fig.width=10}
labels <- sapply(rownames(res_method), extract_genotype)
head(labels)
table(labels)

corr_plot <- res_method %>%
  dplyr::select(where(is.numeric)) %>%
  corrr::correlate() %>%
  shave() %>%
  rplot(print_cor = TRUE) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
print(corr_plot)
ggsave("figures/correlation_plot.png", corr_plot, width = 10, height = 8)
```

## Data cleaning and UMAP (Fig 2E)

```{r umap_all, echo=TRUE, fig.height=6, fig.width=8}
res_method$genotype <- labels

filt_recipe <- recipe(genotype ~ ., data = res_method) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_corr(all_numeric(), threshold = 0.85)
filt_prep <- filt_recipe %>% prep(training = res_method)
tidy(filt_prep, number = 3)
filt_bake <- filt_prep %>% bake(new_data = NULL)

set.seed(179)
umap_res <- umap(filt_bake %>% dplyr::select(-genotype),
                 n_neighbors = 100, min_dist = 0.1)
umap_df  <- as.data.frame(umap_res$layout)
umap_df$genotype <- filt_bake$genotype

# --- Fig 2E: Combined UMAP coloured by genotype ---
umap_genotype <- ggplot(umap_df, aes(x = V1, y = V2, col = genotype, alpha = 0.5)) +
  geom_point() +
  scale_color_manual(values = genotype_colors) +
  labs(title = "UMAP by Genotype", x = "UMAP 1", y = "UMAP 2",
       color = "Genotype") +
  theme_void(base_size = 16, base_family = "Helvetica") +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.title = element_text(size = 12),
    legend.text  = element_text(size = 10)
  )
print(umap_genotype)
ggsave("figures/umap_by_genotype.png", umap_genotype, width = 8, height = 6)
```

## Individual genotype UMAPs (Fig 2E panels)

Each genotype highlighted against a grey background of all other cells.

```{r umap_individual, echo=TRUE, fig.height=6, fig.width=8}
set.seed(179)
umap_res <- umap(filt_bake %>% dplyr::select(-genotype),
                 n_neighbors = 100, min_dist = 0.1)
umap_df  <- as.data.frame(umap_res$layout)
umap_df$genotype <- filt_bake$genotype

genotypes <- unique(umap_df$genotype)

for (i in seq_along(genotypes)) {
  geno <- genotypes[i]
  background_data <- umap_df %>% filter(genotype != geno)
  highlight_data  <- umap_df %>% filter(genotype == geno)

  p <- ggplot() +
    geom_point(data = background_data, aes(x = V1, y = V2),
               color = "gray80", alpha = 0.3, size = 1) +
    geom_point(data = highlight_data, aes(x = V1, y = V2),
               color = genotype_colors[geno], alpha = 0.8, size = 1.5) +
    labs(title = paste("UMAP -", geno), x = "UMAP 1", y = "UMAP 2") +
    theme_void(base_size = 16, base_family = "Helvetica") +
    theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
  print(p)

  ggsave(sprintf("figures/umap_by_genotype_%s.png", geno),
         p, width = 8, height = 6)
  ggsave(sprintf("figures/umap_by_genotype_%s.pdf", geno),
         p, width = 8, height = 6)
  cat(sprintf("Done: umap_by_genotype_%s.png\n", geno))
}
```

## K-means clustering

```{r kmeans_elbow, echo=TRUE, fig.height=6, fig.width=10}
k_values <- 1:10
wss_values <- sapply(k_values, function(k) {
  kmeans(umap_df %>% dplyr::select(-genotype), k, nstart = 25)$tot.withinss
})

elbow_plot <- data.frame(k = k_values, wss = wss_values) %>%
  ggplot(aes(x = k, y = wss)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = k_values) +
  labs(title = "Elbow Method", x = "Number of clusters (k)",
       y = "Total within-cluster sum of squares") +
  theme_minimal()
print(elbow_plot)
ggsave("figures/elbow_plot.png", elbow_plot, width = 10, height = 6)
```

```{r silhouette, echo=TRUE, fig.height=5, fig.width=7}
silhouette_scores <- sapply(k_values[-1], function(k) {
  km <- kmeans(umap_df %>% dplyr::select(-genotype), centers = k, nstart = 25)
  ss <- cluster::silhouette(km$cluster, dist(umap_df %>% dplyr::select(-genotype)))
  mean(ss[, 3])
})

silhouette_plot <- data.frame(k = k_values[-1],
                              silhouette = silhouette_scores) %>%
  ggplot(aes(x = k, y = silhouette)) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = k_values[-1]) +
  labs(title = "Silhouette Method", x = "Number of clusters (k)",
       y = "Average silhouette width") +
  theme_minimal()
print(silhouette_plot)
ggsave("figures/silhouette_plot.png", silhouette_plot, width = 7, height = 5)
```

```{r kmeans_final, echo=TRUE, fig.height=6, fig.width=8}
final_kmeans <- kmeans(umap_df %>% dplyr::select(-genotype),
                       centers = 3, nstart = 25)

umap_df$cluster    <- final_kmeans$cluster
res_method$cluster <- final_kmeans$cluster

# Recode clusters to match manuscript labels
recode_map <- c(`1` = 3, `2` = 1, `3` = 2)
umap_df$cluster    <- dplyr::recode(umap_df$cluster, !!!recode_map)
res_method$cluster <- dplyr::recode(res_method$cluster, !!!recode_map)

cluster_colors <- c("1" = "#a6d96a", "2" = "#abd9e9", "3" = "#fdae61")

cluster_plot <- ggplot(umap_df, aes(x = V1, y = V2,
                                     col = as.factor(cluster))) +
  geom_point(size = 1.5, alpha = 0.8) +
  scale_color_manual(values = cluster_colors) +
  labs(title = "UMAP by Cluster", x = "UMAP 1", y = "UMAP 2",
       color = "Cluster") +
  theme_void(base_size = 16, base_family = "Helvetica") +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        legend.title = element_text(size = 12),
        legend.text  = element_text(size = 10))
print(cluster_plot)
ggsave("figures/umap_by_cluster.png", cluster_plot, width = 8, height = 6)
ggsave("figures/umap_by_cluster.pdf", cluster_plot, width = 8, height = 6)
```

## Cluster composition by genotype

```{r cluster_barplot, echo=TRUE, fig.height=6, fig.width=14}
cluster_composition <- res_method %>%
  dplyr::select(genotype, cluster) %>%
  group_by(genotype, cluster) %>%
  count() %>%
  ungroup(cluster) %>%
  mutate(prop = n / sum(n))

bar_plot <- ggplot(cluster_composition,
       aes(fill = as.factor(cluster), y = prop,
           x = factor(genotype, levels = genotype_order))) +
  geom_bar(position = "stack", stat = "identity") +
  scale_fill_manual(values = cluster_colors) +
  labs(title = "Cluster composition by genotype",
       x = "Genotype", y = "Proportion of cells", fill = "Cluster") +
  theme_minimal()
print(bar_plot)
ggsave("figures/cluster_composition_by_genotype.png", bar_plot,
       width = 14, height = 6)
ggsave("figures/cluster_composition_by_genotype.pdf", bar_plot,
       width = 8, height = 6)
```

## Feature boxplots by cluster

```{r feature_boxplots, echo=TRUE, fig.height=10, fig.width=12}
umap_features  <- setdiff(colnames(filt_bake), "genotype")
df_clustervals <- res_method[c(umap_features, "cluster")]

boxplot_features <- df_clustervals %>%
  pivot_longer(cols = -cluster, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = as.factor(cluster), y = value)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~ feature, scales = "free") +
  labs(x = "Cluster", y = "Feature Value")
print(boxplot_features)
ggsave("figures/features_by_cluster_boxplots.png", boxplot_features,
       width = 12, height = 10)
```

## Statistical tests (Kruskal-Wallis + effect size)

```{r stat_tests, echo=TRUE}
find_meaningful_features <- function(df, p_threshold = 0.05,
                                     effectsize_threshold = 0.1) {
  results <- data.frame(
    feature = colnames(df)[-ncol(df)],
    p_value = sapply(df[-ncol(df)], function(x) {
      kruskal.test(x ~ cluster, data = df)$p.value
    }),
    effectsize = sapply(df[-ncol(df)], function(x) {
      H <- kruskal.test(x ~ cluster, data = df)$statistic
      n <- length(x)
      unname(H / ((n^2 - 1) / (n + 1)))
    })
  )
  results$adj_p <- p.adjust(results$p_value, "fdr")
  results %>%
    filter(adj_p < p_threshold, effectsize > effectsize_threshold) %>%
    arrange(adj_p)
}
significant_features <- find_meaningful_features(df_clustervals)
significant_features
```

## Pairwise cluster tests (Wilcoxon)

```{r pairwise_tests, echo=TRUE}
pairwise_cluster_tests <- function(df, features) {
  results <- list()
  for (feat in features) {
    comparisons <- combn(unique(df$cluster), 2, simplify = FALSE)
    pair_results <- lapply(comparisons, function(pair) {
      test <- wilcox.test(df[df$cluster == pair[1], feat],
                          df[df$cluster == pair[2], feat])
      data.frame(
        feature     = feat,
        cluster1    = pair[1],
        cluster2    = pair[2],
        p_value     = test$p.value,
        median_diff = median(df[df$cluster == pair[1], feat]) -
                      median(df[df$cluster == pair[2], feat])
      )
    })
    results[[feat]] <- bind_rows(pair_results)
  }
  all_results       <- bind_rows(results)
  all_results$adj_p <- p.adjust(all_results$p_value, "fdr")
  all_results
}
pairwise_results <- pairwise_cluster_tests(df_clustervals,
                                           significant_features$feature)
pairwise_results
```

## Cluster-specific features

```{r cluster_specific, echo=TRUE}
find_cluster_specific_features <- function(pairwise_results) {
  pairwise_results %>%
    group_by(feature) %>%
    mutate(
      n_sig = sum(adj_p < 0.001),
      specificity = case_when(
        all(median_diff[adj_p < 0.001] > 0) ~
          paste0("specific_to_", first(cluster1)),
        all(median_diff[adj_p < 0.001] < 0) ~
          paste0("specific_to_", first(cluster2)),
        TRUE ~ "non_specific"
      )
    ) %>%
    filter(specificity != "non_specific") %>%
    arrange(specificity, adj_p)
}
specific_features <- find_cluster_specific_features(pairwise_results)
specific_features
```

## Session info

```{r session_info, echo=TRUE}
sessionInfo()
```
